RED WINE Analysis Part1 by Terry Troutman
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
#library(plyr)
#library(dplyr)

library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(car)
library(reshape)
library(bitops)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("C:/R files")
wine <-read.csv("wineQualityReds.csv")

```

# Overview of the Wine Data

  For more information, read [Cortez et al., 2009].

###Input variables (based on physicochemical tests):
   
   1 - fixed acidity (tartaric acid - g / dm^3)
   
   2 - volatile acidity (acetic acid - g / dm^3)
   
   3 - citric acid (g / dm^3)
   
   4 - residual sugar (g / dm^3)
   
   5 - chlorides (sodium chloride - g / dm^3
   
   6 - free sulfur dioxide (mg / dm^3)
   
   7 - total sulfur dioxide (mg / dm^3)
   
   8 - density (g / cm^3)
   
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   
   11 - alcohol (% by volume)
   
###Output variable (based on sensory data): 
   
   12 - quality (score between 0 and 10)



###Description of attributes:

   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

   2 - volatile acidity: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

   3 - citric acid: found in small quantities, citric acid can add 'freshness' and flavor to wines

   4 - residual sugar: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

   5 - chlorides: the amount of salt in the wine

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

   8 - density: the density of water is close to that of water depending on the percent alcohol and sugar content

   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

   11 - alcohol: the percent alcohol content of the wine

###Output variable (based on sensory data):

   12 - quality (score between 0 and 10)
   
####Structure of Wine Data Set

```{r echo=FALSE}
str(wine)
```

####Summary of the Wine Data Set
After looking at this summary we can note that citric acid is the only variable that contains a zero value.  It is not an outlier there many samples where the measured citric acid content is zero.
  Ignoring the range variable X and quality, it appears that many of the quatnitative measurements have outliers.  This is as defined traditionally where outliers < 1st Quartile - 1.5 *IQR or > 3rd Quartile + 1.5 * IQR.  
  Alchohol, pH and density have outliers but they are not nearly as extreme as the other quantitative variables.

```{r echo=FALSE}
summary(wine)

```



   
#Derived Variables
  There is likely to be a relationship between residual.sugar and acid.  The more residual.sugar is present, the more acid is required to balance it.  So it makes sense to have a variable with SugarToAcid. SugarToAcid is a the ratio of residual.sugar variable to the fixed.acidity variable.  The volatile.acidity measurement is an indicator of spoilage so it was not considered.
  The balance of salt to sugar is also likely to have an effect on the taste of wine.  So another derivied varialble will be SugarToSalt(Chlorides).
  
  
```{r Residual Sugar To Fixed Acidity and  Sugar to Salt}
wine$SugarToAcid <- wine$residual.sugar / wine$fixed.acidity
wine$SugarToSalt <- wine$residual.sugar / wine$chlorides

```
```{r Summary SugarToAcid, echo=FALSE}

summary(wine$SugarToAcid)

```
```{r summary SugarToSalt, echo=FALSE}

summary(wine$SugarToSalt)

```


# Univariate Plots Section


## Histograms

```{r echo=FALSE, Fixed_Acidity}
ggplot(aes(x=fixed.acidity,),data=wine)+geom_histogram(binwidth=0.3) +
  ggtitle("Fixed Acidity")
```
    
The fixed acidity has an alost normal distribution.  If the binwidth gets below 0.3 the variance increases. 
  
```{r echo=FALSE, Volatile_Acidty}
ggplot(aes(x=volatile.acidity),data=wine)+geom_histogram(binwidth=0.05)+
  scale_x_continuous(breaks=seq(0,1.6,0.1))+
  ggtitle("Volatile Acidity")

```
  
The binwidth must be decreased 0.05 seems a reasonable choice.  This data is somewhat bimodal with peak counts around .38 and .6.  There are some outliers on the right.

 
```{r echo=FALSE, Citric_Acid}

ggplot(aes(x=citric.acid),data=wine)+geom_histogram(binwidth=.025)+
  ggtitle("Citric Acid")

```
  
There are many wines with zero citric.acid content.  Most wines have a very low citric.acid content.  Disribution adds no insight.


```{r echo=FALSE}

ggplot(aes(x=residual.sugar),data=wine)+geom_histogram(binwidth=0.25)+
  ggtitle("Residual Sugar")

``` 
  Residual sugar has a somewhat normal distribution with a tail.

```{r echo=FALSE}

ggplot(aes(x=chlorides),data=wine)+geom_histogram(binwidth=.002)+
  scale_x_continuous(breaks=seq(0,.65,.05)) +
  ggtitle("Chlorides")

```
  
Chlorides have a narrow but regular distriubion. There are some outlier on the high end above about .125.
    
```{r echo=FALSE}

ggplot(aes(x=free.sulfur.dioxide),data=wine)+geom_histogram(binwidth=2)+
 ggtitle("Free SO2")
```
   
This is a skewed distribution. Most values are below 40.


```{r echo=FALSE}
  
ggplot(aes(x=total.sulfur.dioxide),data=wine)+geom_histogram(binwidth=2) +
  ggtitle("Total Sulfur Dioxide")

```
  Total Sulfur Dioxide is a skewed distribution.  Most value are below 100.
  The graph resembles and exponential decay.  


```{r echo=FALSE}
ggplot(aes(x=sulphates),data=wine)+geom_histogram(binwidth=.05)+
  ggtitle("Sulphates")

```

This distribution looks normal on the RHS but very steep on the LHS.  
Nearly all values are under 1.0
  
```{r echo=FALSE}
ggplot(aes(x=density),data=wine) + geom_histogram() +
  ggtitle("Density")

```
  This is an almost normal distribution.


```{r echo=FALSE}
ggplot(aes(x=pH),data=wine)+geom_histogram(binwidth=.05) +
  ggtitle("pH")

```

  This is an almost normal distribution.
  
```{r echo=FALSE}

ggplot(aes(x=quality), data=wine)+ geom_histogram()+ggtitle("Quality")

ggplot(aes(x=quality), data=wine)+ geom_histogram()+
  scale_y_log10()+
  ggtitle("Distribution of quality is normalized when a log is taken")

```

  Taking the log of quality makes the distriubtion normal.
  
```{r echo=FALSE}

ggplot(aes(x=SugarToAcid), data=wine)+ geom_histogram()+
  ggtitle("SugarToAcid")

ggplot(aes(x=SugarToAcid), data=wine)+ geom_histogram()+
  scale_x_log10()+
  ggtitle("Sugar to Acid with log10 of x")

  ggplot(aes(x=SugarToSalt),data=wine)+ geom_histogram()+
  ggtitle("SugarToSalt")

ggplot(aes(x=SugarToSalt),data=wine)+ geom_histogram()+
  scale_x_log10()+
  ggtitle("Sugar to Acid with log10 of x")

```

  

    





# Univariate Analysis


No variables besides citric acid have a non-zero value. 
This is also a good sanity check.  For example:
A zero value for pH would indicate that the log of the acid conentration would be infinity.  Wine in its creation produces alcohol; an alcohol content of zero would be very suspicious.   
All wines have some chloride content.  This is interpreted as salt content.  Wine has many compounds but this is generally taken to be NaCl content.
Sulphates and Sulfur dioxide appears to be in this entire wine sampling.
Not all the acid in wine is from citric acid.  Other acidic compounds are present.

### What is the structure of your dataset?

It is a dataframe of 1599 observations(rows) of 13 variables.  Since there are only 1599 rows, it isn't necessary to sample the data. The first variable is the index.  The next 11 variables are quantitative measurements of wine samples.  We will attempt to use the 11 input variables, or a subset of the 11 input variables to establish a model to predict wine quality.  We may also derive variables from the input varialbes.  


### What is/are the main feature(s) of interest in your dataset?

All of the wine is acidic as indicated by a pH <7.  
The median alcohol content of 10.20 is low compared to today's red wine market which I would estimate to be around 12.  This is inherent to the data set.  It may very well be quite typical of red grapes of the vinho verde grape family of Portugal.  

There are some outliers in this data set. The residual sugar Max is almost 6 times the 3rd Quartile.  That value should be discarded. 

Having no factor variables in the input variables makes determining relationships of input variables to output variable more difficult.  Perhaps we can make new factors using the cut function.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?


### Did you create any new variables from existing variables in the dataset?

Yes.  Please see section above titled Derived Varialbes.  I created a new variable of the ratio of residual.sugar to fixed.acidity and another new variable of the ratio of residual.sugar to chlorides.  

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Yes some of the distributions were long tailed so in the graphs I took a log10 of the x scale.  There are outliers but I am hesistant to crop the data.
